Comparison among doublet predictions on benchmarking datasets

### 2024-06-17 analyses/doublet-detection/exploratory-notebooks/03_compare-benchmark-results.Rmd
Use cxds threshold of 0.5 and use scale_color_manual instead of scale_color_identity

- The ground truth calls themselves to which we are comparing consensus calls may not be entirely accurate, since they were also computationally identified generally with demultiplexing algorithms. 
- `cxds`, as we are using it, does not have a specific threshold for calling droplets. 
By contrast, both `scrublet` and `scDblFinder` identify a threshold based on the given dataset they are processing. 
This notebook used a doublet threshold of `>=0.75` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).

@@

theme_set(theme_bw())
```

@@

plot_pca_metrics <- function(df, color_column, false_colors) {
  # Plot PCs colored by performance metric, showing false calls on top
  # false_colors is a vector of colors used for fn and fp points
  # df is expected to contain columns PC1, PC2, and `color_column`. This should _not_ be provided as a string.
  ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(
    size = 0.75, 
    alpha = 0.6
  ) + 
  geom_point(
    data = dplyr::filter(df, {{color_column}} %in% false_colors), 
    size = 0.75
  ) +
  scale_color_identity() +
  theme(legend.position = "none")
}
```

@@

      scdbl_df <- scdbl_tsv |>
        readr::read_tsv(show_col_types = FALSE) |>
        dplyr::select(
          barcodes,
          cxds_score, 
          scdbl_score = score, 
          scdbl_prediction  = class
        ) |>
        # add cxds calls at 0.75 threshold
        dplyr::mutate(
          cxds_prediction = dplyr::if_else(
            cxds_score >= 0.75,
            "doublet",
            "singlet"
          )
        )

@@

      # grab ground truth and PCA coordinates
      sce <- readr::read_rds(sce_file)
      sce_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
        tibble::rownames_to_column(var = "barcodes") |>
        dplyr::select(barcodes,
                      ground_truth = ground_truth_doublets, 
                      PC1 = PCA.1, 
                      PC2 = PCA.2) |> 
        dplyr::left_join(
          scrub_df, 
          by = "barcodes"
        ) |>
        dplyr::left_join(
          scdbl_df, 
          by = "barcodes"
        )

      # Add a consensus call column
      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        dplyr::mutate(consensus_call = dplyr::if_else(
          all(
            c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"
          ),
          "doublet", 
          "singlet"
        )) |>
        dplyr::mutate(
          call_type = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          ), 
          # set associated plotting colors
          call_type_color = dplyr::case_when(
            call_type == "tp" ~ unname(confusion_colors["tp"]),
            call_type == "tn" ~ unname(confusion_colors["tn"]),
            call_type == "fp" ~ unname(confusion_colors["fp"]),
            call_type == "fn" ~ unname(confusion_colors["fn"])
          )
          )

@@

```{r, fig.width = 12, fig.height = 6}
# Make a legend for the confusion-colored PCA
legend_plot <- data.frame(
  x = factor(names(confusion_colors), levels = names(confusion_colors)), y = 1:4
) |>
 ggplot(aes(x = x, y = y, color = x)) + 
  geom_point(size = 3) + 
  scale_color_manual(name = "Metric", values = confusion_colors) 
confusion_legend <- ggpubr::get_legend(legend_plot) |> ggpubr::as_ggplot()

doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {

@@

      # Third, call type
      p3 <- plot_pca_metrics(
        df,
        call_type_color,
        false_colors = unname(c(confusion_colors["fn"], confusion_colors["fp"]))
      )

      # combine and plot
      plot( p1 + p2 + p3 + confusion_legend + plot_annotation(glue::glue("PCA for {dataset}")) + plot_layout(ncol=4, widths = c(1,1,1,0.25)) )
    }
  )
```

>>>>

- The ground truth calls themselves to which we are comparing consensus calls may not be entirely accurate, since they were also computationally identified generally with demultiplexing algorithms. 
- `cxds`, as we are using it, does not have a specific threshold for calling droplets. 
By contrast, both `scrublet` and `scDblFinder` identify a threshold based on the given dataset they are processing. 
This notebook used a doublet threshold of `>=0.5` for `cxds`, which may not be universally suitable (though choosing a universally suitable threshold is not easy either!).

@@

theme_set(theme_bw())

# define threshold used to call cxds
cxds_threshold <- 0.5
```

@@

plot_pca_metrics <- function(df, color_column, metric_colors) {
  # Plot PCs colored by performance metric, showing false calls on top
  # metric_colors is a named vector of colors used for coloring tp/tn/fp/fn
  # df is expected to contain columns PC1, PC2, and `color_column`. This should _not_ be provided as a string.
  ggplot(df) + 
    aes(x = PC1, 
        y = PC2, 
        color = {{color_column}}) +
  geom_point(
    size = 0.75, 
    alpha = 0.6
  ) + 
  geom_point(
    data = dplyr::filter(df, {{color_column}} %in% c("fp", "fn")), 
    size = 0.75
  ) +
  scale_color_manual(name = "Call type", values = metric_colors) +
  theme(legend.position = "bottom")
}
```

@@

      scdbl_df <- scdbl_tsv |>
        readr::read_tsv(show_col_types = FALSE) |>
        dplyr::select(
          barcodes,
          cxds_score, 
          scdbl_score = score, 
          scdbl_prediction  = class
        ) |>
        # add cxds calls at `cxds_threshold` threshold
        dplyr::mutate(
          cxds_prediction = dplyr::if_else(
            cxds_score >= cxds_threshold,
            "doublet",
            "singlet"
          )
        )

@@

      # grab ground truth and PCA coordinates
      sce <- readr::read_rds(sce_file)
      dataset_df <- scuttle::makePerCellDF(sce, use.dimred = "PCA") |>
        tibble::rownames_to_column(var = "barcodes") |>
        dplyr::select(barcodes,
                      ground_truth = ground_truth_doublets, 
                      PC1 = PCA.1, 
                      PC2 = PCA.2) |> 
        dplyr::left_join(
          scrub_df, 
          by = "barcodes"
        ) |>
        dplyr::left_join(
          scdbl_df, 
          by = "barcodes"
        )

      # Add a consensus call column
      dataset_df <- dataset_df |>
        dplyr::rowwise() |>
        dplyr::mutate(consensus_call = dplyr::if_else(
          all(
            c(scdbl_prediction, scrublet_prediction, cxds_prediction) == "doublet"
          ),
          "doublet", 
          "singlet"
        )) |>
        dplyr::mutate(
          call_type = dplyr::case_when(
            consensus_call == "doublet" && ground_truth == "doublet" ~ "tp",
            consensus_call == "singlet" && ground_truth == "singlet" ~ "tn",
            consensus_call == "doublet" && ground_truth == "singlet" ~ "fp",
            consensus_call == "singlet" && ground_truth == "doublet" ~ "fn"
          )
        )

@@

```{r, fig.width = 12, fig.height = 6}
doublet_df_list |>
  purrr::iwalk(
    \(df, dataset) {

@@

      # Third, call type
      p3 <- plot_pca_metrics(
        df,
        call_type, 
        metric_colors = confusion_colors
      )

      # combine and plot
      plot( p1 + p2 + p3 + plot_annotation(glue::glue("PCA for {dataset}")) + plot_layout(ncol=4, widths = c(1,1,1)) )
    }
  )
```
